A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-04-21 jacobian_calc_CMAME.tex
Switched to using latexdiff to track changes, makes things much cleaner

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%%\usepackage{graphics}
%% or use the graphicx package for more complicated commands
%%\usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}
%%\usepackage{color}
%%\usepackage{tkz-base}
\usepackage{pgfplots}
\usepackage{siunitx}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{array}
\usepackage{hyperref}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}

%Package for adding notes/comments/and tracking changes
\usepackage[inline]{trackchanges}
\addeditor{JTF}
\addeditor{MDB}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

@@

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
\author{Michael D. Brothers}
\author{John T. Foster\corref{cor1}}
\ead{john.foster@utsa.edu}
\cortext[cor1]{Corresponding Author}

\author{Harry R. Millwater\corref{}}
\address{Mechanical Engineering Department, The University of Texas at San Antonio}

%% \ead{email address}
%% \ead[url]{home page}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

@@

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

@@

In order to maintain the quadratic convergence properties of Newton's method in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. A goal of the study described in this paper was to establish the suitability of an underexplored method for numerical computation of tangent-stiffness operators, referred to as ``complex-step'', and compare the \note{Re: I.1} \remove{new} method with other techniques for numerical derivative calculation: automatic differentiation, forward finite difference, and central finite difference. The complex-step method was \note{Re: I.1} \remove{newly} implemented in a massively parallel computational peridynamics code for the purpose of this comparison.  The methods were compared through profiling of the code for accuracy, speed, efficiency, and parallel scalability. The research provides data that can serve as practical guide for code developers and analysts faced with choosing which method best suits the needs of their application code.  Additionally, motivated by the reproducible research movement, all of the code, examples, and workflow to regenerate the data and figures in this paper are provided as open source.

@@

In order to maintain the quadratic convergence properties of the first-order Newton's method \cite{belytschko1999nonlinear} \cite[Ch.~13]{young2009} in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent\footnote{\note{Re: II.1} In the context of plasticity thoery, the algorithmic consistency of the tangent-stiffness matrix refers to consistency with the integration of a plasticity model under the Khun-Tucker constraints or similar where the tangent-stiffness matrix is computed using the elastoplastic modulii that are determined from the integration \cite{simo1998}.} tangent-stiffness matrices. For an extremely small class of nonlinear material models, these consistent tangent-stiffness operators can be derived analytically; however, most often in practice, they are found through numerical approximation of derivatives.

A goal of this study was to develop and evaluate a \note{Re: I.1} \change{new }{relatively unexplored}, accurate, and practical method for calculating tangent-stiffness matrices against established methods.  This \note{Re: I.1} \remove{new} method is based on a complex number Taylor series expansion and referred to as CTSE or the ``complex-step'' method. The practical accuracy of CTSE is determined by comparing its measured accuracy relative to the measured accuracy of popular finite differencing techniques used for computing tangent-stiffness matrices. For this comparison the,  exact to machine precision, derivatives computed with \emph{automatic differentiation} serve as the standard.

@@

The scope of the application component of this study was limited to a single material model, an elastic peridynamic solid, implemented in the computational peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific application provided a practical framework for implementing the \note{Re: I.1} \change{new }{complex-step} method and solving engineering problems to generate the data needed to compare the methods. In particular, \emph{Peridigm} was chosen because it combined several helpful characteristics: the utilization of Newton's method and tangent-stiffness matrices for solving nonlinear quasi-static problems, prior inclusion of forward-difference, central-difference, and automatic-differentiation methods needed for comparison to the complex-step method implementation, and being an agile-components code that makes use of distributed computing data structures via Trilinos \cite{trilinos} for efficient parallelization on large clusters. Agile-components is a software development term meaning to structure a software project for continuous development to respond to evolving specifications by making much use of techniques from object-oriented programming and reusing proven pre-existing code to reduce development time and increase reliability. The modularity of \emph{Peridigm} allowed the adding of new features, such as those required by the study.

The aim of this paper is not only to \note{Re: I.1} \change{introduce the new}{study the} complex-step method in the context of evaluating tangent-stiffness matrices, but to serve as a review for other differentiation techniques useful for solving non-linear systems with Newton's method. After presenting background information on the underlying methods to be discussed, presented in this paper are: In Section~\ref{subsec:TS} a description of tangent-stiffness matrices, detailed directions for producing tangent-stiffness matrices with each of the methods identified above, and a description and justification of the \note{Re: I.1} \remove{new} complex-step method for calculating tangent-stiffness matrices, in Section~\ref{subsubsec:Impl} a description of implementing the complex-step method in the \emph{Peridigm} software, in Section~\ref{JGAM} a description of the quantities-of-interest used to rank the methods, in Section~\ref{sec:Res} a presentation and analysis of the results of the comparative study, and in section~\ref{sec:Conc} conclusions and thoughts on potential future work.  Finally, in the interest of replicable research, the reader is referred to the corresponding author's website for \emph{Peridigm} C++ source-code and data necessary to reproduce the results presented here. Additionally, included on the website is an example serial C++ library with classes for solving non-linear systems using the techniques discussed here, as well as two validation and verification example problems which show how to use the library.  The purpose of the serial C++ library is to demonstrate in a less complex manner than in the \emph{Peridigm} code how one would use automatic-differentiation or complex-step to better encourage their use and discussion. \note{Re: II.5} \add{This open source resource is available at the corresponding author's website or at} \href{https://github.com/utsa-idl/Complex_Step_Jacobian_Project_Archive}{https://github.com/utsa-idl/Complex\_Step\_Jacobian\_Project\_Archive}.

@@

It is possible to approximate derivatives quite accurately with a technique based on a complex-variable Taylor series expansion of a function.  This method was first described by Lyness and Moler \cite{lyness1967numerical,lyness1968differentiation} and has more recently been rediscovered for use in engineering analysis.  The basic idea is that a model parameter can be made complex and expanded in a Taylor series about a small perturbation along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align} 
    f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i h}{1!}  \notag \\ &+ \frac{\partial^2 f(x)}{\partial x^2} \frac{(i h)^2}{2!} + \frac{\partial^3 f(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots.
\label{eqn:complexTaylor} 
\end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor}) and solving for the first derivative and ignoring terms of $\mathcal{O}\left(h^2 \right)$ yields an estimate
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{\mbox{Im} \left( f (x + i h) \right)}{h} +\mathcal{O}\left( h^2 \right).  
\label{eqn:complexFirstDeriv} 
\end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only utilizing one functional evaluation of a perturbed model parameter along the imaginary axis.  The step-size $h$ is arbitrary and can be made as small as practical (even to machine precision without the dangers of roundoff error as in finite differences) to yield an accurate estimate of this derivative.  The only disadvantage of this technique is the necessity of requiring the functions to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a simple example that illustrates the use of the complex-step method for the derivative calculation of a function. 
%
\note{Re: II.6} \add{For completeness, the forward- and central-difference formulas corresponding to equation}~(\ref{eqn:complexFirstDeriv}) \add{are} \cite[Chap. 4.1.3]{chapra2010}
%
\begin{equation} 
\frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x)}{h} +\mathcal{O}\left( h \right), 
\label{eqn:forwardFirstDeriv} 
\end{equation}
%
\begin{equation} 
\frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x - h)}{2h} +\mathcal{O}\left( h^2 \right). 
\label{eqn:centeredFirstDeriv} 
\end{equation} 
%
\add{The order of the truncation error in the forward-difference formula (equation}~(\ref{eqn:forwardFirstDeriv})\add{) indicates that for a given $h$ this method is less accurate an approximation than either central-difference or complex-step which have errors of $\mathcal{O}(h^2)$.}

\note{Re: I.4} \add{As mentioned earlier, the CS method was originally introduced in a paper by Lyness and Moler} \cite{lyness1967numerical}\add{.  A follow-on contribution by Lyness demonstrated a ``truncated'' approximate version of the original CS method to make it suitable for use on digital computers} \cite{lyness1968differentiation}\add{.  After these original papers, little can be found in the literature regarding the method until Squire and Trapp showed through numerical experiments that CS offers superior accuracy to FD and CD} \cite{squire1998using} \add{in computing numerical derivatives. Papers by Newman et al.\ } \cite{newman1998} \add{and Anderson et al.\ } \cite{anderson2001sensitivity} \add{discuss computing sensitivities in a CFD analysis using CS and compares the results to sensitivities produced with central-difference derivatives. The work is also notable for identifying automatic-differentiation as a competing method to CS and suggesting that complex-step could be used to compute Jacobians of residuals for use in Newton-Krylov schemes for solving non-linear systems. Perez-Foguet et al.\ } \cite{perez2000numerical, perez2012numerical} \add{used the CS method to compute algorithmically consistent elastoplastic moduli in non-linear solid mechanics problems.  A paper by Joaquim et al.\ } \cite{martins2000automated} \add{investigates using CS for computing aerodynamic sensitivities. Additionally, higher-order accuracy versions of CS are discussed, as is the task of implementation in most widely used general purpose programming languages. Discussed is the relative ease of implementing a CS scheme versus an automatic-differentiation based scheme for computing sensitivities in an established code.  
Martins et al.\ } \cite{martins2003complex}  \add{demonstrates that the CS method can be performed by using complex datatypes for the variables of functions comprising an analysis code. The accuracy of this approach, with respect to the derivatives themselves, is compared to automatic-differentiation in a serial analysis code. Lai and Crassidis} \cite{lai2008extensions} \add{discuss first- and second-order CS methods for computing derivatives and choosing optimal step sizes.  Al-Mohy and Higham} \cite{al2010complex} \add{investigate the use of CS in computing Fr\'{e}chet derivatives of matrix functions. Jin et al.\ } \cite{jin2010improved} \add{discuss using CS to compute sensitivities in a solid mechanics FEM analysis.  Voorhees et al.\ } \cite{voorhees2011complex} \add{discusses the background of the CS method and frames it as a subset of Fourier Differentiation. CS and Fourier Differentiation are then applied to computing sensitivities within a solid mechanics finite element analysis. They implemented the complex-values as additional degrees of freedom in the finite element stiffness matrices using a real-valued matrix representation of the complex-numbers where required. In a follow-on work,  Millwater et al.\ } \cite{millwater2013application} \add{extended this concept, calling it ZFEM, and implemented the method into user defined element types for use with the popular ABAQUS} \cite{systemes2012abaqus} \add{FEA software package.  Nishida et al.\ } \cite{nishida2013} \add{used the complex-step method to compute Jacobian matrices for implementation in a Levenberg-Marquardt optimization routine for design-of-experiments purposes.  Finally, Abreu et al.\ } \cite{abreu2013generalization} \add{discusses computing derivatives with a version of the CS method which features the combined use of real and imaginary valued step components, as in Fourier differentiation.  They suggests that the modified CS method can achieve extended approximation accuracy up to fourth order.  This recent literature has introduced the application of CS to single-dimension and multidimensional derivatives, sensitivity analysis, non-linear solution schemes, optimization, the extension of CS for higher accuracy, and implementation strategies for CS in new and existing codes.  The current paper presents a comparison of the complex-step method to the other methods for numerical differentiation in as close to a one-to-one implementation as possible for direct comparison purposes in a large-scale engineering analysis code. To the author's knowledge it is the first application of the complex-step method to solve peridynamics mechanics problems, and the first implementation and comparison of the method in large-scale distributed memory parallel setting. }

@@

\note[]{Re: I.5, next section is taken from the thesis.}
\section{Simple example of CS, AD and FD for solution of nonlinear problems.}
\label{sec:WorkingExamples}

@@

\note[]{Re: I.2, John I think you would better handle expanding this section,
I was going to essentially copy silling:psa.}\change[]{Briefly, peridynamics}Peridynamics \cite{silling2000ret,silling:psa,silling2010peridynamic}
is a nonlocal reformulation of the partial-differential equations that provide
the statement of momentum balance in classical continuum mechanics. Its primary
goal is to avoid the use of spatial derivatives in the balance equations or
constitutive laws such that discontinuous displacements, i.e. cracks, are
mathematically consistent with the governing equations. It has demonstrated
great promise in modeling problems with pervasive failure
\cite{littlewood2010}, interesting phenomenon such as dynamic crack branching
\cite{ha2010sod}, and is being extended to multiphyics phenomena
\cite{bobaru2011peridynamic,katiyar2013} and the problem of multi-scale
coupling to molecular simulations \cite{seleson2009peridynamics,seleson2014atom}.

@@

\note[]{Re: I.2. John, this was to be the skeleton for the new sections on peridynamics.}
\subsubsection{Vector states for state based peridynamic theory}
Because the vector state, which is fair to say an uncommon notation, is used in the literature discussing state based peridynamic theory
it will be enumerated here for the sake of continuity. 
%
Paraphrasing, let a spherical region, centered about a
location in threespace, called a horizon or neighborhood with a certain radius
be defined. A vector state of order $m$ is a mapping of a vector contained
within the horizon to a tensor of order $m$ \cite{silling:psa}. Vector states
have have their own mathematical properties and definitions for sums, composition,
point product, reference state, identity scalar state, null reference state, null scalar state,
dot product, magnitude state, unit state, and direction state which are detailed in \cite{silling:psa}. 
It is important to think of vector states as non-linear, not necessarily continuous maps of vectors to real Euclidian space, where second order tensors are linear maps of three dimensional vectors to threespace. 
%
One important tool mechanics gains from the concept of vector states is a way
to represent complicated deformations within a body in the form of a state
function. Whereas a symmetric second order tensor maps a spherical region in a
body in the reference configuration to an ellipsoid in the deformed
configuration, a vector state can map a spherical region in a body into a
complicated shape \cite{silling:psa}.

@@

where "$\mathcal{H}_{\boldsymbol{x}}$ is a spherical neighborhood" with a
radius, centered at the original coordinates of a peridynamic node,
$\boldsymbol{x}$, and $\boldsymbol{\underline{T}}$ is the "force vector state
field". The force vector state field maps vectors in the original coordinate system 
measured from the peridynamic node at the center of the neighborhood or 'horizon' to
neighboring peridynamic nodes that fall within the radius of the horizon. $\boldsymbol{b}$ denotes a body force density. Equation~\ref{eqn:PDMotion} may be simplified based on the assumption that
$\boldsymbol{\underline{T}}$ is Riemann integrable and that integral is uniformly
convergent \cite{silling:psa} as,

@@

\note[]{Re: I.2} \add[]{
\subsubsection{The tangent stiffness matrix in a quasi-static peridynamic simulation}
The tangent-stiffness matrix is the Fr\'{e}chet derivative of the residual wrt the
deformation state.  For the purposes of this work, the residual represents the unbalanced forces in the
static peridynamic momentum equation. The internal force
density is computed in the peridynamic momentum equation by computing the
volume integral over a horizon of the Fr\'{e}chet derivative of the 
strain energy density state wrt to the deformed configuration inner product with the deformation state, per node,
for each of the three directions in the original reference frame.

For the finite difference type methods in \emph{Peridigm}, the perturbation
method is employed by probing the deformed configuration one degree of freedom
at a time and so allowing the derivative of the force density state wrt the
deformation state for all nodes to be computed. 
}

>>>>

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%%\usepackage{graphics}
%% or use the graphicx package for more complicated commands
%%\usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}
%%\usepackage{color}
%%\usepackage{tkz-base}
\usepackage{pgfplots}
\pgfplotsset{compat=1.8}
\usepackage{siunitx}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{array}
\usepackage{hyperref}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}

%Package for adding notes/comments/and tracking changes
\usepackage[]{changes}
\usepackage[]{todonotes}

%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

@@

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \addedress for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
\author{Michael D. Brothers}
\author{John T. Foster\corref{cor1}}
\ead{john.foster@utsa.edu}
\cortext[cor1]{Corresponding Author} 
\author{Harry R. Millwater\corref{}}
\address{Mechanical Engineering Department, The University of Texas at San Antonio}

%% \ead{email address}
%% \ead[url]{home page}
%% \addedress{Address\fnref{label3}}
%% \fntext[label3]{}

@@

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \addedress[label1]{<address>}
%% \addedress[label2]{<address>}

@@

    In order to maintain the quadratic convergence properties of Newton's method in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. A goal of the study described in this paper was to establish the suitability of an underexplored method for numerical computation of tangent-stiffness operators, referred to as ``complex-step'', and compare the \todo[inline]{Re: I.1} method with other techniques for numerical derivative calculation: automatic differentiation, forward finite difference, and central finite difference. The complex-step method was \todo[inline]{Re: I.1} implemented in a massively parallel computational peridynamics code for the purpose of this comparison.  The methods were compared through profiling of the code for accuracy, speed, efficiency, and parallel scalability. The research provides data that can serve as practical guide for code developers and analysts faced with choosing which method best suits the needs of their application code.  Additionally, motivated by the reproducible research movement, all of the code, examples, and workflow to regenerate the data and figures in this paper are provided as open source.

@@

In order to maintain the quadratic convergence properties of the first-order Newton's method \cite{belytschko1999nonlinear} \cite[Ch.~13]{young2009} in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent\todo[disable]{Re: II.1}\footnote{In the context of plasticity thoery, the algorithmic consistency of the tangent-stiffness matrix refers to consistency with the integration of a plasticity model under the Khun-Tucker constraints or similar where the tangent-stiffness matrix is computed using the elastoplastic modulii that are determined from the integration \cite{simo1998}.} tangent-stiffness matrices. For an extremely small class of nonlinear material models, these consistent tangent-stiffness operators can be derived analytically; however, most often in practice, they are found through numerical approximation of derivatives.

A goal of this study was to develop and evaluate a \todo[disable]{Re: I.1} relatively unexplored, accurate, and practical method for calculating tangent-stiffness matrices against established methods.  This  \todo[disable]{Re: I.1} method is based on a complex number Taylor series expansion and referred to as CTSE or the ``complex-step'' method. The practical accuracy of CTSE is determined by comparing its measured accuracy relative to the measured accuracy of popular finite differencing techniques used for computing tangent-stiffness matrices. For this comparison the,  exact to machine precision, derivatives computed with \emph{automatic differentiation} serve as the standard.

@@

The scope of the application component of this study was limited to a single material model, an elastic peridynamic solid, implemented in the computational peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific application provided a practical framework for implementing the  \todo[disable]{Re: I.1} complex-step method and solving engineering problems to generate the data needed to compare the methods. In particular, \emph{Peridigm} was chosen because it combined several helpful characteristics: the utilization of Newton's method and tangent-stiffness matrices for solving nonlinear quasi-static problems, prior inclusion of forward-difference, central-difference, and automatic-differentiation methods needed for comparison to the complex-step method implementation, and being an agile-components code that makes use of distributed computing data structures via Trilinos \cite{trilinos} for efficient parallelization on large clusters. Agile-components is a software development term meaning to structure a software project for continuous development to respond to evolving specifications by making much use of techniques from object-oriented programming and reusing proven pre-existing code to reduce development time and increase reliability. The modularity of \emph{Peridigm} allowed the adding of new features, such as those required by the study.

The aim of this paper is not only to \todo[disable]{Re: I.1} study the complex-step method in the context of evaluating tangent-stiffness matrices, but to serve as a review for other differentiation techniques useful for solving non-linear systems with Newton's method. After presenting background information on the underlying methods to be discussed, presented in this paper are: In Section~\ref{subsec:TS} a description of tangent-stiffness matrices, detailed directions for producing tangent-stiffness matrices with each of the methods identified above, and a description and justification of the  \todo[disable]{Re: I.1} complex-step method for calculating tangent-stiffness matrices, in Section~\ref{subsubsec:Impl} a description of implementing the complex-step method in the \emph{Peridigm} software, in Section~\ref{JGAM} a description of the quantities-of-interest used to rank the methods, in Section~\ref{sec:Res} a presentation and analysis of the results of the comparative study, and in section~\ref{sec:Conc} conclusions and thoughts on potential future work.  Finally, in the interest of replicable research, the reader is referred to the corresponding author's website for \emph{Peridigm} C++ source-code and data necessary to reproduce the results presented here. Additionally, included on the website is an example serial C++ library with classes for solving non-linear systems using the techniques discussed here, as well as two validation and verification example problems which show how to use the library.  The purpose of the serial C++ library is to demonstrate in a less complex manner than in the \emph{Peridigm} code how one would use automatic-differentiation or complex-step to better encourage their use and discussion.  \todo[disable]{Re: II.5}This open source resource is available at the corresponding author's website or at \href{https://github.com/utsa-idl/Complex_Step_Jacobian_Project_Archive}{https://github.com/utsa-idl/Complex\_Step\_Jacobian\_Project\_Archive}.

@@

It is possible to approximate derivatives quite accurately with a technique based on a complex-variable Taylor series expansion of a function.  This method was first described by Lyness and Moler \cite{lyness1967numerical,lyness1968differentiation} and has more recently been rediscovered for use in engineering analysis.  The basic idea is that a model parameter can be made complex and expanded in a Taylor series about a small perturbation along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align} 
    f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i h}{1!}  \notag \\ &+ \frac{\partial^2 f(x)}{\partial x^2} \frac{(i h)^2}{2!} + \frac{\partial^3 f(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots.
\label{eqn:complexTaylor} 
\end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor}) and solving for the first derivative and ignoring terms of $\mathcal{O}\left(h^2 \right)$ yields an estimate
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{\mbox{Im} \left( f (x + i h) \right)}{h} +\mathcal{O}\left( h^2 \right).  
\label{eqn:complexFirstDeriv} 
\end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only utilizing one functional evaluation of a perturbed model parameter along the imaginary axis.  The step-size $h$ is arbitrary and can be made as small as practical (even to machine precision without the dangers of roundoff error as in finite differences) to yield an accurate estimate of this derivative.  The only disadvantage of this technique is the necessity of requiring the functions to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a simple example that illustrates the use of the complex-step method for the derivative calculation of a function. 
%
\todo[disable]{Re: II.6}For completeness, the forward- and central-difference formulas corresponding to equation~(\ref{eqn:complexFirstDeriv}) are \cite[Chap. 4.1.3]{chapra2010}
%
\begin{equation} 
\frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x)}{h} +\mathcal{O}\left( h \right), 
\label{eqn:forwardFirstDeriv} 
\end{equation}
%
\begin{equation} 
\frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x - h)}{2h} +\mathcal{O}\left( h^2 \right). 
\label{eqn:centeredFirstDeriv} 
\end{equation} 
%
The order of the truncation error in the forward-difference formula (equation~(\ref{eqn:forwardFirstDeriv})) indicates that for a given $h$ this method is less accurate an approximation than either central-difference or complex-step which have errors of $\mathcal{O}(h^2)$.

\todo[disable]{Re: I.4}As mentioned earlier, the CS method was originally introduced in a paper by Lyness and Moler \cite{lyness1967numerical}.  A follow-on contribution by Lyness demonstrated a ``truncated'' approximate version of the original CS method to make it suitable for use on digital computers \cite{lyness1968differentiation}.  After these original papers, little can be found in the literature regarding the method until Squire and Trapp showed through numerical experiments that CS offers superior accuracy to FD and CD \cite{squire1998using} in computing numerical derivatives. Papers by Newman et al.\ \cite{newman1998} and Anderson et al.\ \cite{anderson2001sensitivity} discuss computing sensitivities in a CFD analysis using CS and compares the results to sensitivities produced with central-difference derivatives. The work is also notable for identifying automatic-differentiation as a competing method to CS and suggesting that complex-step could be used to compute Jacobians of residuals for use in Newton-Krylov schemes for solving non-linear systems. Perez-Foguet et al.\ \cite{perez2000numerical, perez2012numerical} used the CS method to compute algorithmically consistent elastoplastic moduli in non-linear solid mechanics problems.  A paper by Joaquim et al.\ \cite{martins2000automated} investigates using CS for computing aerodynamic sensitivities. Additionally, higher-order accuracy versions of CS are discussed, as is the task of implementation in most widely used general purpose programming languages. Discussed is the relative ease of implementing a CS scheme versus an automatic-differentiation based scheme for computing sensitivities in an established code.

Martins et al.\ \cite{martins2003complex} demonstrates that the CS method can be performed by using complex datatypes for the variables of functions comprising an analysis code. The accuracy of this approach, with respect to the derivatives themselves, is compared to automatic-differentiation in a serial analysis code. Lai and Crassidis \cite{lai2008extensions} discuss first- and second-order CS methods for computing derivatives and choosing optimal step sizes.  Al-Mohy and Higham \cite{al2010complex} investigate the use of CS in computing Fr\'{e}chet derivatives of matrix functions. Jin et al.\ \cite{jin2010improved} discuss using CS to compute sensitivities in a solid mechanics FEM analysis.  Voorhees et al.\ \cite{voorhees2011complex} discusses the background of the CS method and frames it as a subset of Fourier Differentiation. CS and Fourier Differentiation are then applied to computing sensitivities within a solid mechanics finite element analysis. They implemented the complex-values as additional degrees of freedom in the finite element stiffness matrices using a real-valued matrix representation of the complex-numbers where required. In a follow-on work,  Millwater et al.\  \cite{millwater2013application} extended this concept, calling it ZFEM, and implemented the method into user defined element types for use with the popular ABAQUS \cite{systemes2012abaqus} FEA software package.  Nishida et al.\ \cite{nishida2013} used the complex-step method to compute Jacobian matrices for implementation in a Levenberg-Marquardt optimization routine for design-of-experiments purposes.  Finally, Abreu et al.\ \cite{abreu2013generalization} discusses computing derivatives with a version of the CS method which features the combined use of real and imaginary valued step components, as in Fourier differentiation.  They suggests that the modified CS method can achieve extended approximation accuracy up to fourth order.  This recent literature has introduced the application of CS to single-dimension and multidimensional derivatives, sensitivity analysis, non-linear solution schemes, optimization, the extension of CS for higher accuracy, and implementation strategies for CS in new and existing codes.  The current paper presents a comparison of the complex-step method to the other methods for numerical differentiation in as close to a one-to-one implementation as possible for direct comparison purposes in a large-scale engineering analysis code. To the author's knowledge it is the first application of the complex-step method to solve peridynamics mechanics problems, and the first implementation and comparison of the method in large-scale distributed memory parallel setting.

@@

\todo[disable][]{Re: I.5}
\section{Simple example of CS, AD and FD for solution of nonlinear problems.}
\label{sec:WorkingExamples}

@@

\todo[disable]{Re: I.2} Peridynamics \cite{silling2000ret,silling:psa,silling2010peridynamic}
is a nonlocal reformulation of the partial-differential equations that provide
the statement of momentum balance in classical continuum mechanics. Its primary
goal is to avoid the use of spatial derivatives in the balance equations or
constitutive laws such that discontinuous displacements, i.e. cracks, are
mathematically consistent with the governing equations. It has demonstrated
great promise in modeling problems with pervasive failure
\cite{littlewood2010}, interesting phenomenon such as dynamic crack branching
\cite{ha2010sod}, and is being extended to multiphyics phenomena
\cite{bobaru2011peridynamic,katiyar2013} and the problem of multi-scale
coupling to molecular simulations \cite{seleson2009peridynamics,seleson2014atom}.

@@

\subsubsection{Vector states for state based peridynamic theory}
Because the vector state, which is fair to say an uncommon notation, is used in the literature discussing state based peridynamic theory
it will be enumerated here for the sake of continuity. 
%
Paraphrasing, let a spherical region, centered about a
location in threespace, called a horizon or neighborhood with a certain radius
be defined. A vector state of order $m$ is a mapping of a vector contained
within the horizon to a tensor of order $m$ \cite{silling:psa}. Vector states
have have their own mathematical properties and definitions for sums, composition,
point product, reference state, identity scalar state, null reference state, null scalar state,
dot product, magnitude state, unit state, and direction state which are detailed in \cite{silling:psa}. 
It is important to think of vector states as non-linear, not necessarily continuous maps of vectors to real Euclidian space, where second order tensors are linear maps of three dimensional vectors to threespace. 
%
One important tool mechanics gains from the concept of vector states is a way
to represent complicated deformations within a body in the form of a state
function. Whereas a symmetric second order tensor maps a spherical region in a
body in the reference configuration to an ellipsoid in the deformed
configuration, a vector state can map a spherical region in a body into a
complicated shape \cite{silling:psa}.

@@

where "$\mathcal{H}_{\boldsymbol{x}}$ is a spherical neighborhood" with a
radius, centered at the original coordinates of a peridynamic node,
$\boldsymbol{x}$, and $\boldsymbol{\underline{T}}$ is the "force vector state
field". The force vector state field maps vectors in the original coordinate system 
measured from the peridynamic node at the center of the neighborhood or `horizon' to
neighboring peridynamic nodes that fall within the radius of the horizon. $\boldsymbol{b}$ denotes a body force density. Equation~(\ref{eqn:PDMotion}) may be simplified based on the assumption that $\boldsymbol{\underline{T}}$ is Riemann integrable and that integral is uniformly
convergent \cite{silling:psa} as,

@@

\todo[disable]{Re: I.2}
\subsubsection{The tangent stiffness matrix in a quasi-static peridynamic simulation}
The tangent-stiffness matrix is the Fr\'{e}chet derivative of the residual wrt the
deformation state.  For the purposes of this work, the residual represents the unbalanced forces in the
static peridynamic momentum equation. The internal force
density is computed in the peridynamic momentum equation by computing the
volume integral over a horizon of the Fr\'{e}chet derivative of the 
strain energy density state wrt to the deformed configuration inner product with the deformation state, per node,
for each of the three directions in the original reference frame.

For the finite difference type methods in \emph{Peridigm}, the perturbation
method is employed by probing the deformed configuration one degree of freedom
at a time and so allowing the derivative of the force density state wrt the
deformation state for all nodes to be computed.

\subsubsection{Implementing the complex-step method in \emph{Peridigm}} 
\label{subsubsec:Impl}
%
The CS method was implemented in \emph{Peridigm} in a manner that follows
closely the implementation of the FD technique.  However, because of
\emph{Peridigm}'s reliance on \emph{Trilinos} and specifically the
\emph{Epetra} package, special steps had to be followed to allow the use of
complex number data types.  This is because \emph{Epetra} vectors, which can be
thought of as distributed memory parallel cousins to standard C++
Standard Template Library vectors, are hard coded to be
of type {\tt double} and can not be simply declared as having a complex data
type. \todo[disable]{Re: I.3}For the reader's convenience, \emph{Epetra} vectors and STL vectors are what are known as 'container classes', whose purpose in object oriented computer programming is to
    automate the creation, accessing, modification and destruction of the
    underlying contained data. Templated container classes allow for user
    choice in what properties that underlying data exhibits.  For example, in
    this work, complex arithmetic cannot be performed on the data contained
    within an \emph{Epetra} vector because that data, of type {\tt double} does
    not have the property of following the rules of complex arithmetic. The
    reason the non-templated \emph{Epetra} vector could not be dispensed with
    for the purposes of this work is because the \emph{Epetra} vector was
tightly integrated with \emph{Peridigm's} parallel communication strategy and
linear algebra code. Program methods that were used in the course of applying
the FD method were copied, renamed and re-written to follow the CS formula.
Because of restrictions imposed by the container class being used, this
involved changing the data type of intermediate variables which were locally
scoped, and dynamically casting persistent memory variables to complex data
types where necessary. \todo[disable]{Re: I.3}For the reader's
convenience, 'locally scoped' in the C++ programming language refers to
data which is created within a context more specific than the context of
the entire program, or 'translation unit'.  Most data created this way is
destroyed when that specific context is left. The topic of contexts in C++
is very broad, so for the purposes of this work context refers to a program
method or function that begins, concludes and yields a result, also
destroying locally scoped data. Dynamic casting for the purposes of this
work refers to copying the value of data contained within an \emph{Epetra}
vector into locally scoped data that has the property of following the rules of
complex arithmetic, where a mapping called 'type casting' allows the
translation of data between different types. No attempt was made to ``tune''
the CS code in any way for performance. The algorithm  was implemented in such
as way that it was in one-to-one correspondence with the FD and CD methods;
however, it must be noted that the overhead associated with the dynamic casts
in the CS method likely hurts its overall performance and this issue could be
alleviated by using the next generation of templated \emph{TPetra} vectors
available in \emph{Trilinos} that are capable of taking an explicit
instantiation of any data type including {\tt complex}. \todo[disable][]{Re:
I.3}For the purposes of this work, an explicit instantiation
refers to requesting the creation of an object, or instance, of a templated
class or container class that holds data which has the property of
following desired mathematic or logic rules. The version of \emph{Peridigm}
modified to implement CS can be found in the paper repository.
